Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2021-10-15 doc/JFLA2022/paper.tex
fill in proof sizes where possible

In 2020, members of \community transferred the \gaia source code to GitHub and adapted it for recent releases of the Mathematical Components library, which \gaia heavily relies on. Anonymous volunteers (``collaborators of Nicolas Bourbaki'') then finished the only in-progress proof left by Grimm. At around 165,000 lines of code, \gaia is currently one of the largest maintained open source \coq projects~\cite{Gaia}.

\gaia contains definitions of ordinals in Cantor and Veblen normal form~\cite{grimm:hal-00911710}, adapted from the historical Cantor contribution~\cite{CantorContrib}. The data types for ordinals are essentially defined the same way in Gaia and \Hydras, but they are not identical inside \coq, e.g., due to residing in different modules. Moreover, there are small differences in the implementation of ordinal arithmetic due to different evolutionary paths taken since divergence from the ancestral library.
%
Nevertheless, we were able to establish an initial correspondence between both implementations of ordinals through rewriting lemmas. For instance, we proved that multiplication of ordinals in Cantor normal form in \Hydras refine Gaia's corresponding multiplication operation, and were able to import Gaia's proof of associativity of the multiplication almost for free.

@@

Both due to its prior stages~\cite{PCiota,JFLA2018paper} and the recent integrations described just above, the current \Hydras Coq code has a complex inheritance. Figure~\ref{fig:genealogy} illustrates the relationships between \Hydras packages and show their ancestry from historical Coq contributions. To indicate the scope and relative sizes of packages, Table~\ref{tbl:loc} breaks down lines of code for each \Hydras package, as reported by the \texttt{coqwc} tool. We also name and link to the repository in \community for each package.

@@

\draw[->] (equations) -- (hydras) ;
\draw[->] (mathcomp) -- (gaia) ;
\draw[->] (mathcomp) -- (chains) ;
\draw[->] (paramcoq.south) -- (chains) ;
\draw[->] (pockcc) -- (goedelcc) ;
\end{tikzpicture}
}
\caption{Genealogy and dependencies for \Hydras packages. Dotted boxes represent historical Coq contributions, while regular boxes represent maintained Coq packages. Dark gray packages are maintained in the \Hydras GitHub repository, while light gray packages are maintained in other \community repositories. Dotted lines represent Coq code ancestry, while regular lines represent direct code dependencies.}
  \label{fig:genealogy}
\end{figure}

@@

\begin{table}[ht]
\centering
\footnotesize
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\textbf{Package name} & \textbf{Repository} & \textbf{Package on opam} & \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
\hline
Pocklington & \href{https://github.com/coq-community/pocklington}{\texttt{pocklington}} & \texttt{coq-pocklington} & \texttt{8.12.0} & & \\
\hline
Goedel & \href{https://github.com/coq-community/goedel}{\texttt{goedel}} & \texttt{coq-goedel} & \texttt{8.13.0} & & \\
\hline
Gaia & \href{https://github.com/coq-community/gaia}{\texttt{gaia}} & \texttt{coq-gaia} & \texttt{1.12} & & \\
\hline
Hydra-battles & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-hydra-battles} & \texttt{0.5} & & \\
\hline
Addition-chains & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-addition-chains} & \texttt{0.5} & & \\
\hline
Gaia-hydras & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-gaia-hydras} & \texttt{0.5} & & \\
\hline
\end{tabular}
\caption{Current numbers of lines of code for \Hydras packages.}
\label{tbl:loc}
\end{table}

@@

\Hydras wants to be provide a connection between scientific literature (\emph{e.g.},~\cite{KP82, KS81, schutte}) and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the \coq learner, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. %It may be also used as a library, through opam packages.
It is also a medium sized library (more than 50K LOC), dependent on various tools and libraries of the \coq ecosystem, which  may be also used to experiment new techniques of continuous maintenance of the code and its documentation.

>>>>

In 2020, members of \community transferred the \gaia source code to GitHub and adapted it for recent releases of the Mathematical Components library, which \gaia heavily relies on. Anonymous volunteers (``collaborators of Nicolas Bourbaki'') then finished the only in-progress proof left by Grimm. At around 155,000 lines of code, \gaia is currently one of the largest maintained open source \coq projects~\cite{Gaia}.

\gaia contains definitions of ordinals in Cantor and Veblen normal form~\cite{grimm:hal-00911710}, adapted from the historical Cantor contribution~\cite{CantorContrib}. The data types for ordinals are essentially defined the same way in \gaia and \Hydras, but they are not identical inside \coq, e.g., due to residing in different modules. Moreover, there are small differences in the implementation of ordinal arithmetic due to different evolutionary paths taken since divergence from the ancestral library.
%
Nevertheless, we were able to establish an initial correspondence between both implementations of ordinals through rewriting lemmas. For instance, we proved that multiplication of ordinals in Cantor normal form in \Hydras refine \gaia's corresponding multiplication operation, and were able to import \gaia's proof of associativity of the multiplication almost for free.

@@

Both due to its prior stages~\cite{PCiota,JFLA2018paper} and the recent integrations described just above, the current \Hydras \coq code has a complex inheritance. Figure~\ref{fig:genealogy} illustrates the relationships between \Hydras packages and show their ancestry from historical Coq contributions. To indicate the scope and relative sizes of packages, Table~\ref{tbl:loc} breaks down lines of code for each \Hydras package, as reported by the \texttt{coqwc} tool, and lists their repository names in \community and identifiers in the \coq opam package index~\cite{CoqPackageIndex}.

@@

\draw[->] (equations) -- (hydras) ;
\draw[->] (mathcomp) -- (gaia) ;
\draw[->] (mathcomp) -- (chains) ;
\draw[->] (paramcoq.south) -- (chains) ;
\draw[->] (pockcc) -- (goedelcc) ;
\end{tikzpicture}
}
\caption{Genealogy and dependencies for \Hydras packages. Dotted boxes represent historical Coq contributions, while regular boxes represent maintained \coq packages. Dark gray packages are maintained in the \Hydras GitHub repository, while light gray packages are maintained in other \community repositories. Dotted lines represent \coq code ancestry, while regular lines represent direct code dependencies.}
  \label{fig:genealogy}
\end{figure}

@@

\begin{table}[ht]
\centering
\footnotesize
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\textbf{Package name} & \textbf{Repository} & \textbf{Index identifier} & \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
\hline
Pocklington & \href{https://github.com/coq-community/pocklington}{\texttt{pocklington}} & \texttt{coq-pocklington} & \texttt{8.12.0} & 825 & 3,798 \\
\hline
Goedel & \href{https://github.com/coq-community/goedel}{\texttt{goedel}} & \texttt{coq-goedel} & \texttt{8.13.0} & 2,799 & 10,762 \\
\hline
Gaia & \href{https://github.com/coq-community/gaia}{\texttt{gaia}} & \texttt{coq-gaia} & \texttt{1.12} & 28,850 & 124,839 \\
\hline
Hydra-battles & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-hydra-battles} & \texttt{0.5} & & \\
\hline
Addition-chains & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-addition-chains} & \texttt{0.5} & & \\
\hline
Gaia-hydras & \href{https://github.com/coq-community/hydra-battles}{\texttt{hydra-battles}} & \texttt{coq-gaia-hydras} & \texttt{0.5} & & \\
\hline
\end{tabular}
\caption{Current numbers of lines of code for \Hydras packages.}
\label{tbl:loc}
\end{table}

@@

\Hydras wants to be provide a connection between scientific literature (\emph{e.g.},~\cite{KP82, KS81, schutte}) and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the \coq learner, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. %It may be also used as a library, through opam packages.
It is also a medium sized library (more than 50,000 LOC), dependent on various tools and libraries of the \coq ecosystem, which  may be also used to experiment new techniques of continuous maintenance of the code and its documentation.
